### 2017-12-10 paper/jfp-submission/response.txt

We thank the reviewers for their detailed comments and suggestions.

As requested, we have added a paragraph at the end of the Introduction
that describes the major changes from the ICFP paper.

In the rest of this document, we will respond to the larger comments and
concerns of the reviewers, and describe the changes we have made to
address them.

# Questions about the Evaluation

## User Study

@@

Our use of IRR is intended to account for potential bias in the
raters. One such form of bias (conscious or not) could be bias in favor
of NanoMaLy since it is new and interesting, and the raters (even if
they are not directly connected to the project) may want to see it
succeed. Due to the graphical nature of NanoMaLy's traces it is not easy
to hide whether a given answer was assisted by OCaml or NanoMaLy from
the raters. A further form of potential bias is inconsistency in
evaluating the student answers, which were long-form, English
explanations of the error/fix. There is a surprising amount of
subjectivity involved in judging, based on a couple sentences, whether
the student truly understood the error.

For these reasons we feel that our use of IRR is justified to lend
credibility to the results.

(We have not added this discussion to the paper as the IRR test is
really meant to be a preliminary "sanity-check", but would be happy to
if the reviewers feel it would be helpful.)

## Investigating the "safe" programs

Reviewer 1
> page 23, regarding section "witness exists", I got puzzled. The study
> there is on 50 programs. It was found that 2% of the programs admit a
> witness. I deduce this must correspond to one single program. But
> then, the next sentence tells that "slightly over half of these
> programs involved had some particular property, while the rest had
> some other property."  There semms to be some inconcistency there.

# Questions about limitations

## Traversal Bias

Reviewer 2
> 1. you do not address the issue of bias, the fact that locations
> towards the end of the program are more likely to be crashed at. This
> is similar to the left-to-right-bias of algorithm W, although in your
> case the left-to-rightness is in the trace, not the syntax tree (as
> addressed for that case in work by Hage and Heeren (VODCA 2008) and
> the work of Erwig and Chen).  Does your implementation allow general
> datatypes so that I can have case with many arms?

Indeed, this was an oversight on our part. We have added a new paragraph
to section 5.8 to discuss our bias.

## Other language features

Reviewer 1
> The tool is restricted to a small subset of OCaml, in particular it does
> not support side-effects. While many FP courses begin by focusing on
> pure programs only, there are some intringuing type errors that students
> face only when working with effectful programs, e.g. forgetting to apply
> a function to the unit argument. I speculate that it would be /not/ so
> easy to generalize the present work to deal with side effects, because
> of the need to maintain the state and report it to the user.

While side effects would certainly make the visualization more complex,
we don't believe they present a fundamental limitation to our approach.
Your example of the student forgetting to supply the unit argument is
actually quite similar to the problem posed by curried functions. Either
the function will be consumed where it's return value was expected,
which would produce an error similar to our `wwhile` example, our it
will be discarded entirely as the call was intended solely for its side
effect. In the latter case, we could simply add a restriction that any
such discarded values must be unit (it is commonly considered a best
practice to explicitly discard the results of functions called just for
their side effects with a `void` function).

Reviewer 2
> 1. explain how things would be if applied to other functional
> languages, in particular Haskell. What about laziness? What about
> overloaded numerals?  It seems to me that these are quite
> problematic. For example, the 0 in your fac program will be Num a =>
> a.

Overloading is indeed problematic for us. We discuss this in the new
section 5.3 (under the heading "Ad-Hoc Polymorphsim"), both describing
the issue and presenting a possible solution. We return to the issue in
section 5.8 and mention how a language like Haskell with pervasive
overloading would exacerbate the problem.

## Scalability

Reviewer 1
> Random-test generation would likely fail to produce counter-examples
> on slightly more complex programs such as red-black trees; for such
> programs, more advanced techniques such as constraint-based test
> generation are probably required. I was not able to quite understand
> whether the approach presented in the paper could be extended to
> accomodate constraint-based test generation.

Indeed, we discuss the limitations of random generation in section
5.8. We have added a sentence there describing how we could replace
random generation with constraint-based generation.

Reviewer 2
> 5. your approximation of functions to fun is rather strong, and it is
> unclear to me how well that works. I guess since all executions start
> from the beginning of the programs keeps everything rather concrete,
> so that you do not need to invent functions. But I am not sure, and I
> would like to see a thorough discussion. I am fine with novice
> programs being short and simple since that is quite likely to be the
> case when they resort to Nanomaly, but when you do that, it should be
> clear that you do.  What may be related to this is the idea to deal
> with Currying. What about functions like foldr of which the result is
> a type variable? I know that some static work on error diagnosis
> sometimes runs into a problem of some sort and since you work at the
> dynamic level this may not be the case for you. But do say and explain
> so.

### 2017-12-10 paper/jfp-submission/evaluation.tex
more response

There are, of course, drawbacks to our approach. Four that stand out
are:
%
(1) coverage limits due to random generation;
%
(2) dealing with explosions in the size of generated traces;
%
%(3) the inability to handle certain instances of infinite types; and
(3) our use of a non-parametric function type; and
%
(4) handling ad-hoc polymorphism.

\paragraph{Random Generation}
Random test generation has difficulty generating highly constrained
values, \eg\ red-black trees or a pair of equal integers. If the type
error is hidden behind a complex branch condition \nanomaly\ may not be
able to trigger it. Exhaustive testing and dynamic-symbolic execution
can address this short-coming by performing an exhaustive search for
inputs (\emph{resp}.\ paths through the program). As our experiments
show, however, novice programs do not appear to require more advanced
search techniques, likely because they tend to be simple.

@@

>>>>

There are, of course, drawbacks to our approach. Five that stand out
are:
%
(1) coverage limits due to random generation;
%
(2) dealing with explosions in the size of generated traces;
%
%(3) the inability to handle certain instances of infinite types; and
(3) our use of a non-parametric function type;
%
(4) handling ad-hoc polymorphism; and
%
(5) traversal bias.

\paragraph{Random Generation}
Random test generation has difficulty generating highly constrained
values, \eg\ red-black trees or a pair of equal integers. If the type
error is hidden behind a complex branch condition \nanomaly\ may not be
able to trigger it. Exhaustive testing and dynamic-symbolic execution
can address this short-coming by performing an exhaustive search for
inputs (\emph{resp}.\ paths through the program). Our approach does not
rely on random generation, we could easily substitute it for
dynamic-symbolic execution by extending the evaluation relation to
maintain a path condition and replacing the \gensym function with a call
to the constraint solver.
%
As our experiments show, however, novice programs do not appear to
require more advanced search techniques, likely because they tend to be
simple.

@@

\paragraph{Traversal Bias}

A common problem with typecheckers is that the order in which the
typechecker traverses the abstract syntax tree \emph{biases} it in favor
of blaming expressions that are seen later~\cite{McAdam1998-ub}. This
usually takes the form of a left-to-right bias with respect to the
source code (terms that appear later \emph{textually} are more likely to
be blamed), but in our case the bias is with respect to the execution
trace. Incorporating our notion of type sources from
\S~\ref{sec:locating} into the visualization, \eg by including those
reductions in the initial visualization, may help alleviate our bias.
% in a similar manner to the proposals of McAdam~\shortcite{McAdam1998-ub},

% Several solutions to the bias problem have been proposed.
% %
% Hage and Heeren~\shortcite{Hage2009-em} describe a system that allows
% the compiler author to selectively control the bias and thus produce
% better errors, while McAdam~\shortcite{McAdam1998-ub},
